---
title: "Bayesian Updating for Surveillance – Hypothetical Plant Pathogen"
author: "Jacobo Robledo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: FALSE
    df_print: paged
  word_document:
    toc: true
    toc_depth: '2'
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 7,
  fig.height = 4
)
```

# 1 Problem statement

Regulatory authorities are evaluating whether to expand a quarantine zone to include 100 hectares of non-commercial host plants. The expansion will proceed **if**

$$
\Pr\bigl(\pi > 0.01 \mid \text{data}\bigr) \;\ge\; 0.80,
$$

where

-   $\pi$ = true fraction of infected host plants in the surveyed region;
-   “data’’ = annual surveillance results (positives $k_t$ out of constant samples $n_t = 350$).

# 2 Bayesian model

## 2.1 Likelihood – Binomial sampling

$k_t \mid \pi \;\sim\; \text{Binomial}\!\bigl(n_t,\pi\bigr), \qquad n_t = 350.$

## 2.2 Prior – expert elicitation

Experts provided a prior disease occurrence of:

$$
\pi \;\sim\; \text{Beta}(\alpha_0 = 1,\; \beta_0 = 999).
$$

*This reflects one “imaginary” positive and 999 negatives.*

Updating rule (Beta–Binomial conjugacy):

$$
\boxed{\;
\alpha_t = \alpha_{t-1} + k_t,\quad
\beta_t  = \beta_{t-1}  + (n_t - k_t)
\;}
$$

## 2.3 Posterior summaries

-   Mean $= \alpha_t / (\alpha_t + \beta_t)$
-   95 % CrI = $[q_{0.025},\,q_{0.975}]$ with $q_p = \mathrm{qbeta}(p,\alpha_t,\beta_t)$
-   Tail probability $\Pr(\pi>0.01)=1-\mathrm{pbeta}(0.01,\alpha_t,\beta_t)$

# 3 Surveillance data

```{r data, echo = FALSE}
posit   <- c(0, 5, 10, 20)
samples <- rep(350, 4)
knitr::kable(
  data.frame(Year = 1:4, Samples = samples, Positives = posit),
  caption = "Annual surveillance data (350 samples per year)"
)
```

# 4 Bayesian update in R

```{r posterior-computation}
# Prior parameters
alpha0 <- 1
beta0  <- 999

# Surveillance data
samples <- rep(350, 4)
posit   <- c(0, 5, 10, 20)

# Posterior parameters
alpha <- c(alpha0, alpha0 + cumsum(posit))
beta  <- c(beta0,  beta0  + cumsum(samples - posit))

# Posterior summaries
post_mean <- alpha / (alpha + beta)
ci_low    <- qbeta(0.025, alpha, beta)
ci_high   <- qbeta(0.975, alpha, beta)
prob_gt1  <- 1 - pbeta(0.01, alpha, beta)

results <- data.frame(
  Stage   = c("Prior", paste("Year", 1:4)),
  Alpha   = alpha,
  Beta    = beta,
  MeanPct = round(100 * post_mean, 3),
  Low95   = round(100 * ci_low, 3),
  High95  = round(100 * ci_high, 3),
  Pr_gt1  = round(100 * prob_gt1, 2)
)

knitr::kable(
  results,
  caption = "Posterior mean, 95 % CrI, and Pr(π > 1 %)"
)
```

# 5 Visualising priors and posteriors

```{r plot}
library(ggplot2)
library(reshape2)

uf_cols <- c("#0021A5", "#375EA5", "#6E90C1", "#B56648", "#FA4616")

x <- seq(0, 0.03, length.out = 4000)
dens <- sapply(seq_along(alpha), function(i) dbeta(x, alpha[i], beta[i]))
colnames(dens) <- results$Stage
dens_df  <- cbind.data.frame(x, dens)
dens_long <- melt(dens_df, id.vars = "x",
                  variable.name = "Stage", value.name = "Density")

ggplot() +
  geom_area(
    data = subset(dens_long, x >= 0.01),
    aes(x, Density, fill = Stage),
    position = "identity",
    alpha = 0.25, colour = NA
  ) +
  geom_line(
    data = dens_long,
    aes(x, Density, colour = Stage),
    size = 1.1, alpha = 0.8
  ) +
  geom_vline(xintercept = 0.01, linetype = "dashed") +
  annotate("text",
           x = 0.0105, y = max(dens_long$Density),
           label = "1 % policy threshold", hjust = 0, vjust = 1.4, size = 3) +
  scale_colour_manual(values = uf_cols, name = "Distribution") +
  scale_fill_manual(values = uf_cols,  name = "Shaded area\nπ > 1 %") +
  labs(title = "Disease prevalence in non-commercial hosts\nPrior and yearly posteriors",
       x = "Prevalence (fraction of hosts infected)",
       y = "Density") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

# 6 Interpretation

-   **Year 1** (0 / 350): mean 0.074 %; Pr(π \> 1 %) ≈ 0 %.
-   **Year 3** (10 / 350 added): mean 0.78 %; Pr(π \> 1 %) ≈ 13 %.
-   **Year 4** (20 / 350 added): mean 1.50 %; Pr(π \> 1 %) ≈ 99 %. The policy threshold is surpassed; expansion of quarantine is statistically supported.

# 7 Technical appendix

## 7.1 Beta–Binomial conjugacy

$$
p(\pi \mid k) \;\propto\; \pi^{\alpha_0+k-1}(1-\pi)^{\beta_0+n-k-1}
\;=\; \text{Beta}\bigl(\alpha_0+k,\; \beta_0+n-k\bigr).
$$

## 7.2 Exact tail probability

$$
\Pr(\pi > 0.01) = 1 - F_{\text{Beta}}(0.01;\alpha,\beta)
               = 1 - \text{pbeta}(0.01,\alpha,\beta).
$$


### Get edidatble figures and tables
```{r, include = TRUE}
# Install if needed: install.packages("writexl")
library(writexl)

write_xlsx(results, path = "./results.xlsx")

# Install if needed: install.packages("officer"); install.packages("rvg")
library(officer)
library(rvg)

# Recreate the plot
plt <- ggplot() +
  geom_area(
    data = subset(dens_long, x >= 0.01),
    aes(x, Density, fill = Stage),
    position = "identity",
    alpha = 0.25, colour = NA
  ) +
  geom_line(
    data = dens_long,
    aes(x, Density, colour = Stage),
    size = 1.1, alpha = 0.8
  ) +
  geom_vline(xintercept = 0.01, linetype = "dashed") +
  annotate("text",
           x = 0.0105, y = max(dens_long$Density),
           label = "1 % policy threshold", hjust = 0, vjust = 1.4, size = 3) +
  scale_colour_manual(values = uf_cols, name = "Distribution") +
  scale_fill_manual(values = uf_cols,  name = "Shaded area\nπ > 1 %") +
  labs(title = "Disease prevalence in non-commercial hosts\nPrior and yearly posteriors",
       x = "Prevalence (fraction of hosts infected)",
       y = "Density") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")

# Create a PowerPoint slide and insert the plot
doc <- read_pptx()
doc <- add_slide(doc, layout = "Title and Content", master = "Office Theme")
doc <- ph_with(doc, dml(ggobj = plt), location = ph_location_fullsize())

print(doc, target = "plot.pptx")

```